\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-07-22 analysis/paper/paper.Rmd
review new sections

---
documentclass: jss
author:
  - name: Sahir Rai Bhatnagar *
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon *
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. * joint co-authors
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- TRUE
eval_cs4 <- TRUE
eval_colophon <- TRUE

@@

In this article, we present an \proglang{R} package that combines the ideas of Hanley & Miettinen into a simple interface. The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile. Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function. In addition, we provide additional features, including performance metrics and visualization tools.

In what follows, we first recall some theoretical details on case-base sampling and its use for estimating parameteric hazard functions. We then give a short review of existing \proglang{R} packages that implement comparable features as \pkg{casebase}. Next, we provide some details about the implementation of case-base sampling in our package, and we give a brief survey of its main functions. This is followed by four case studies that illustrate the flexibility and capabilities of \pkg{casebase}. We show how the same framework can be used for competing risk analyses, penalized estimation, and for studies with time-dependent exposures. Finally, we end the article with a discussion of the results and of future directions.

@@

To describe the theoretical foundations of case-base sampling, we use the framework of counting processes. In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras; the interested reader can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case]. First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$. For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of non-informative censoring is treated in Saarela [-@saarela2016case]). We assume a continuous time model, which implies that the counting process jumps are less than or equal to one. We are interested in modelling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
$$\lambda_{i}(t) dt = E[dN_{i}(t)].$$

@@

If the hazard function $\lambda_{i}(t; \theta)$ is parametrized in terms of $\theta$, we could define an estimator $\hat{\theta}$ by maximization of the likelihood expression
$$L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^\tau \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)}\left(1 - \lambda_{i}(t;\theta)\right)^{1 - dN_{i}(t)},$$
where $\prod_{t\in[0,u)}$ represents a product integral from $0$ to $u$. However, the integral over time makes the computation and maximisation of $L_0(\theta)$ challenging.

Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual contributions of the form
$$P(dN_{i}(t) \mid dQ_{i}(t) = 1) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
Therefore, we can define an estimating equation for $\theta$ as follows:
$$L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.$$
When a logarithmic link function is used for modelling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$. Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios. Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent. Nonetheless, Saarela [-@saarela2016case] showed that the logarithm of this estimating function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\hat{\theta}$ is asymptotically normally distributed.

@@

Finally, several packages include functions to estimate the CIF. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explictly modeled. Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below). Packages providing this functionality include \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing-risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g. Breslow's estimator) or parametric estimators (e.g. spline functions). Packages that implement this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category often produces estimates in the form of step-wise functions. Providing smooth estimates of the CIF was one of the main motivations for introducting case-base sampling in survival analysis.

@@

The functions in the casebase package can be divided into two categories: 1) exploratory data analysis, in the form of population-time plots; and 2) parametric modeling of the hazard function. We explicitly aimed at being compatible with both \code{data.frame}s and \code{data.table}s. This is evident in some of the coding choices we made, and it is also reflected in our unit tests.

@@

The case-base sampling approach described in Section \ref{theory} can be visualized in the form of a population time plot. They are extremely informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis. We facilitate this task in the \pkg{casebase} package using the \code{popTime} function and \code{plot} method as follows:

@@

The splitting of these tasks gives the user the flexibility to create population-time plots with the graphics system of their choice using the object created in Step 1, while the object created in Step 2 allows one to leverage the \pkg{ggplot2} [@ggplot2] family of functions for further customization.

To illustrate these functions we will use the European Randomized Study of Prostate Cancer Screening (ERSPC) data [@schroder2009screening] which was extracted using the approach described in [@liu2014recovering] and available through the \pkg{casebase} package. This data contains the individual observations for 159,893 men between the ages of 55 and 69 years at entry who were recruited from seven European countries.

We first create the necessary dataset for producing the population time plot using the \code{casebase::popTime} function. In this example we stratify by the treatment group. The resulting object inherits from class `popTime` and stores the exposure variable as an attribute:

@@

We then pass this object to the corresponding \code{plot} method which leverages the \pkg{ggplot2} package to build population time plots. 
```{r plot-stratified-erspc-data-code, eval = FALSE, echo = TRUE}
plot(pt_object, add.base.series = TRUE)
```

```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for ERSPC data."}
plot(pt_object, 
     add.base.series = TRUE,
     ratio = 0.7,
     facet.params = list(ncol = 2),
     ribbon.params = list(fill = "gray50"),
     case.params = list(size = 0.8),
     base.params = list(size = 0.8)) + 
  paper_gg_theme +  
  scale_y_continuous(labels = label_number_si())
```

Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by first adding a layer for the area representing the population time in gray, with subjects having the least amount of observation time plotted at the top of the y-axis. We immediately notice a unique _step shape_ in the population time area which is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of the 4 years 1996 to 1999. This, coupled with the uniform December 31 2006 censoring date, lead to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time (i.e. from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and in particular the January 1 entries in successive years. Tracked forwards in time (i.e. from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes). Since the Swedish and Belgian centres were the last to close their recruitment - in December 2003 - the minimum potential follow-up is three years. Tracked further forwards in time (i.e. after year 3) the attrition is a combination of deaths and staggered entries.

Next, layers for the case series and base series are added. The case series is sampled at random vertically on the plot to avoid having all points along the upper edge of the plot, while the base series is sampled horizontally with sampling weight proportional to their follow-up time. By randomly distributing the points, we can get a better sense of the incidence density. In the ERSPC data, we see that more events are observed later on in time. Therefore a constant hazard model would not be appropriate in this instance as it would overestimate the cumulative incidence earlier on in time, and underestimate it later on. The reader is referred to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population time plot.

@@

  - \code{glm}: This is the familiar logistic regression.
  - \code{glmnet}: This option allows for variable selection using Lasso or elastic-net (cf. Case Study 3). This functionality is provided through the \code{glmnet} package [@friedman2010jss].
  - \code{gam}: This option provides support for *Generalized Additive Models* via the \code{gam} package [@hastie1987generalized].
  - \code{gbm}: This option provides support for *Gradient Boosted Trees* via the \code{gbm} package. This feature is still experimental.

@@

We investigate the differences in survival between the control and screening arms in the ERSPC study presented earlier using casebase sampling to fit a parametric hazard function. We fit four models that differ in the functional form of time; (1) excluded from the linear predictor (Eq.\eqref{eq:exp}), (2) linear function (Eq.\eqref{eq:gomp}), (3) log function (Eq.\eqref{eq:weibull}), and (4) a smooth function using cubic B-splines. The models are fit using \code{fitSmoothHazard}. Since the output object from \code{fitSmoothHazard} inherits from the \code{glm} class, we see a familiar result when using the function \code{summary}:

@@

summary(fits[[2]])
```

```{r, eval = eval_cs1, echo=FALSE}
summary(fits[[2]])
```

The \code{absoluteRisk} function takes as input the \code{fitSmoothHazard} object and returns a matrix where each column corresponds to the covariate profiles specified in the \code{newdata} argument, and each row corresponds to a specified time point:

@@

head(risks[[2]])
```

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from casebase on the Cox model CIF. The CIF estimates for the exponential model in panel (1) overestimate the cumulative incidence earlier on in time, and underestimate it later on. Based on what we observed in the population time plot, where more events are observed later on in time, this poor fit for the exponential hazard is expected. This example demonstrates the benefits of population time plots as an exploratory analysis tool.

In Figure \ref{fig:erspc-cox-cif} panels (2)-(4), we notice a better fit with increasing complexities of time. As noted above, the usual asymptotic results hold for likelihood ratio tests built using casebase sampling models. Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) splines model:

@@

As expected, we see that splines model provides a better fit. Similarly, we can compare the AIC for all four casebase models:

@@

```{r erspc-cox-cif, eval = eval_cs1, echo = FALSE, fig.width=8, fig.height=6, fig.cap="Cumulative incidence functions for control and screening groups in the ERSPC data. In each of the panels, we plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression. (1) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution. (2) Linear function of time which is equivalent to a Gompertz distribution. (3) The natural logarithm of time which is equivalent to a Weibull distribution. (4) Cubic B-spline expansion of time."}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
myCols <- cbPalette[c(4,7)]
new_data <- data.frame(
  ScrArm = c("Control group", "Screening group")#,
  # ignore = 99
)

@@

```{r erspc-casebase-cif, eval = FALSE, echo = FALSE}
# smooth_risk_exp <- casebase::absoluteRisk(
#   object = casebase_exponential,
#   time = seq(0, 15, 0.1),
#   newdata = new_data
# )
# 
# plot(survfit(cox_model, newdata = new_data),
#   xlab = "Years since Randomization",
#   ylab = "Cumulative Incidence (%)",
#   fun = "event",
#   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
#   main = sprintf(
#     "Estimated Cumulative Incidence (risk) of Death from Prostate 
#                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
#     exp(coef(cox_model)),
#     exp(confint(cox_model))[1],
#     exp(confint(cox_model))[2]
#   ),
#   ylim = c(0, 2), yscale = 100
# )
# lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
# lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)
# 
# 
# legend("topleft",
#   legend = c(
#     "Control group (Cox)", "Control group (Casebase)",
#     "Screening group (Cox)", "Screening group (Casebase)"
#   ),
#   col = c("red", "red", "blue", "blue"),
#   lty = c(1, 2, 1, 2),
#   bg = "gray90"
# )
```

@@

      # The maximised log-likelihoods are the same between survreg and flexsurvreg, 
      # however the parameterisation is different: the first coefficient (Intercept) 
      # reported by survreg is log(mu), and survreg's "scale" is
      # dweibull’s (thus flexsurvreg)'s 1 / shape
      res <- -fit[[model]]$res[covariate,] / (1 / fit[[model]]$res["shape",])

@@

In the following table, we present a side-by-side comparison of the hazard ratios and confidence intervals estimated from \code{fitSmoothHazard} and the corresponding parametric model using \code{survival::survreg}, as well as the Cox model estimate. The hazard ratio estimates and confidence intervals are similar across all four models. This reinforces the idea that, under proportional hazards, we do not need to model the full hazard to obtain reliable estimates of the hazard ratio. Of course, different parametric models for the hazards give rise to qualitatively different estimates for the CIF as shown in Figure \ref{fig:erspc-cox-cif}.

```{r print-erspc-estimates, eval = TRUE}
## ---- print-sim-table ----

>>>>

---
documentclass: jss
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. *SRB and MT contributed equally to this work.
keywords:
  # at least one keyword must be supplied
  formatted: [survival analysis, absolute risk, data visualization]
  plain:     [survival analysis, absolute risk, data visualization]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- FALSE
eval_cs4 <- TRUE
eval_colophon <- TRUE

@@

In this article, we present an \proglang{R} package that combines the ideas of Hanley & Miettinen into a simple interface. The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile. Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function. In addition, we provide extensive visualization tools.

In what follows, we first recall some theoretical details on case-base sampling and its use for estimating parametric hazard functions. We then give a short review of existing \proglang{R} packages that implement comparable features as \pkg{casebase}. Next, we provide some details about the implementation of case-base sampling in our package, and we give a brief survey of its main functions. This is followed by four case studies that illustrate the flexibility and capabilities of \pkg{casebase}. We show how the same framework can be used for competing risk analyses, penalized estimation, and for studies with time-dependent exposures. Finally, we end the article with a discussion of the results and of future directions.

@@

To describe the theoretical foundations of case-base sampling, we use the framework of counting processes. In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras; the interested reader can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case]. First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$. For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of non-informative censoring is treated in Saarela [-@saarela2016case]). We assume a continuous time model, which implies that the counting process jumps are less than or equal to one. We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
$$\lambda_{i}(t) dt = E[dN_{i}(t)].$$

@@

If the hazard function $\lambda_{i}(t; \theta)$ is parametrized in terms of $\theta$, we could define an estimator $\hat{\theta}$ by maximization of the likelihood expression
$$L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^\tau \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)}\left(1 - \lambda_{i}(t;\theta)\right)^{1 - dN_{i}(t)},$$
where $\prod_{t\in[0,u)}$ represents a product integral from $0$ to $u$. However, the integral over time makes the computation and maximization of $L_0(\theta)$ challenging.

Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual contributions of the form
$$P(dN_{i}(t) \mid dQ_{i}(t) = 1) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
Therefore, we can define an estimating equation for $\theta$ as follows:
$$L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.$$
When a logarithmic link function is used for modeling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$. Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios. Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent. Nonetheless, Saarela [-@saarela2016case] showed that the logarithm of this estimating function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\hat{\theta}$ is asymptotically normally distributed.

@@

Finally, several packages include functions to estimate the CIF. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled. Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below). Packages providing this functionality include \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing-risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g. Breslow's estimator) or parametric estimators (e.g. spline functions). Packages that implement this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category often produces estimates in the form of step-wise functions. Providing smooth estimates of the CIF was one of the main motivations for introducing case-base sampling in survival analysis.

@@

The functions in the \pkg{casebase} package can be divided into two categories: 1) exploratory data analysis, in the form of population-time plots; and 2) parametric modeling of the hazard function. We strove for compatibility with both \code{data.frame}s and \code{data.table}s; this can be seen in the coding choices we made and the unit tests we wrote.

@@

The case-base sampling approach described in Section \ref{theory} can be visualized in the form of a population time plot. These plots are extremely informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis. The \code{popTime} function and \code{plot} method facilitate this task:

@@

By splitting these tasks, we give flexibility to the user. While the method call in Step 2 allows further customization by using the \pkg{ggplot2} [@ggplot2] family of functions, users can use for the graphics system of their choice to create population-time plots from the object created in Step 1.

To illustrate these functions, we will use the European Randomized Study of Prostate Cancer Screening (ERSPC) data [@schroder2009screening] which was extracted using the approach described in Liu *et al.* [-@liu2014recovering] This dataset is available through the \pkg{casebase} package. It contains the individual observations for 159,893 men  from seven European countries, who were between the ages of 55 and 69 years when recruited for the trial.

We first create the necessary dataset for producing the population time plot using the \code{popTime} function. In this example, we stratify the plot by treatment group. The resulting object inherits from class \code{popTime} and stores the exposure variable as an attribute:

@@

We then pass this object to the corresponding \code{plot} method:

```{r plot-stratified-erspc-data-code, eval = FALSE, echo = TRUE}
plot(pt_object, add.base.series = TRUE)
```

```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for ERSPC data."}
label_number_si <- function (accuracy = 1, unit = NULL, sep = NULL, ...) {
    sep <- if (is.null(unit)) 
        ""
    else " "
    scales:::force_all(accuracy, ...)
    function(x) {
        breaks <- c(0, 10^c(K = 3, M = 6, B = 9, T = 12))
        n_suffix <- cut(abs(x), breaks = c(unname(breaks), Inf), 
            labels = c(names(breaks)), right = FALSE)
        n_suffix[is.na(n_suffix)] <- ""
        suffix <- paste0(sep, n_suffix, unit)
        scale <- 1/breaks[n_suffix]
        scale[which(scale %in% c(Inf, NA))] <- 1
        number(x, accuracy = accuracy, scale = unname(scale), 
            suffix = suffix, ...)
    }
}
plot(pt_object, 
     add.base.series = TRUE,
     ratio = 0.7,
     facet.params = list(ncol = 2),
     ribbon.params = list(fill = "gray50"),
     case.params = list(size = 0.8),
     base.params = list(size = 0.8)) + 
  paper_gg_theme +  
  scale_y_continuous(labels = label_number_si())
```

Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by first adding a layer for the area representing the population time in gray, with subjects having the least amount of observation time plotted at the top of the y-axis. We immediately notice a distinctive *step-wise shape* in the population time area. This is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of year from 1996 to 1999. Coupled with the uniform December 31 2006 censoring date, this led to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time (i.e. from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and the January 1 entries in successive years. Tracked forwards in time (i.e. from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes). Since the Swedish and Belgian centres were the last to complete recruitment--in December 2003--the minimum potential follow-up is three years. Tracked further forwards in time (i.e. after year 3) the attrition is a combination of deaths and staggered entries. As we can see, population-time plots summarise a wealth of information about the study into a simple graph.

Next, layers for the case series and base series are added. The case series is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area. By randomly distributing the cases, we can get a better sense of the incidence density. In Figure \ref{fig:plot-stratified-erspc-data}, we see that more events are observed at later follow-up times. Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the cumulative incidence earlier on in time, and underestimate it later on. Finally, the base series is sampled horizontally with sampling weight proportional to their follow-up time. The reader should refer to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population-time plot.

@@

  - \code{glm}: This is the familiar logistic regression.
  - \code{glmnet}: This option allows for variable selection using Lasso or elastic-net (cf. Case Study 3). This functionality is provided through the \pkg{glmnet} package [@friedman2010jss].
  - \code{gam}: This option provides support for *Generalized Additive Models* via the \pkg{mgcv} package [@hastie1987generalized].
  - \code{gbm}: This option provides support for *Gradient Boosted Trees* via the \pkg{gbm} package. This feature is still experimental.

@@

For our first case study, we return to the ERSPC study and investigate the differences in survival between the control and screening arms. We fit four models that differ in which functional form of time is used: 1) excluded from the linear predictor, 2) linear function, 3) log function, and 4) a smooth function using cubic B-splines. The models are fit using \code{fitSmoothHazard}. Since the output object from \code{fitSmoothHazard} inherits from the \code{glm} class, we can directly use the function \code{summary}:

@@

summary(fits[["gompertz"]])
```

```{r, eval = eval_cs1, echo=FALSE}
summary(fits[["gompertz"]])
```

Next, the \code{absoluteRisk} function takes as input the \code{fitSmoothHazard} object and returns a matrix where each column corresponds to the covariate profiles specified in the \code{newdata} argument, and each row corresponds to a specified time point:

@@

head(risks[["gompertz"]])
```

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from \pkg{casebase} on the Cox model CIF. The CIF estimates for the exponential model in panel (1) overestimate the cumulative incidence earlier on in time, and underestimate it later on. Based on our earlier discussion of the population-time plot, this poor fit for the exponential hazard was expected.

In Figure \ref{fig:erspc-cox-cif} (panels 2--4), we notice a better fit with increasing complexity of our model for time. As noted above, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) splines model:

@@

As expected, we see that splines model provides a better fit. Similarly, we can compare the AIC for all four case-base models:

@@

```{r erspc-cox-cif, eval = eval_cs1, echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIFs for control and screening groups in the ERSPC data. In each of the panels, we plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from the case-base sampling scheme (dashed line) with different functional forms of time. (1) The time variable is excluded (exponential). (2) Linear function of time (Gompertz). (3) The natural logarithm (Weibull). (4) Cubic B-spline expansion of time."}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
myCols <- cbPalette[c(4,7)]
new_data <- data.frame(
  ScrArm = c("Control group", "Screening group")#,
  # ignore = 99
)

@@

```{r erspc-casebase-cif, eval = FALSE, echo = FALSE}
# smooth_risk_exp <- casebase::absoluteRisk(
#   object = casebase_exponential,
#   time = seq(0, 15, 0.1),
#   newdata = new_data
# )
# 
# plot(survfit(cox_model, newdata = new_data),
#   xlab = "Years since Randomization",
#   ylab = "Cumulative Incidence (%)",
#   fun = "event",
#   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
#   main = sprintf(
#     "Estimated Cumulative Incidence (risk) of Death from Prostate 
#                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
#     exp(coef(cox_model)),
#     exp(confint(cox_model))[1],
#     exp(confint(cox_model))[2]
#   ),
#   ylim = c(0, 2), yscale = 100
# )
# lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
# lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)
# 
# legend("topleft",
#   legend = c(
#     "Control group (Cox)", "Control group (Casebase)",
#     "Screening group (Cox)", "Screening group (Casebase)"
#   ),
#   col = c("red", "red", "blue", "blue"),
#   lty = c(1, 2, 1, 2),
#   bg = "gray90"
# )
```

@@

      # The maximized log-likelihoods are the same between survreg and flexsurvreg, 
      # however the parameterisation is different: the first coefficient (Intercept) 
      # reported by survreg is log(mu), and survreg's "scale" is
      # dweibull's (thus flexsurvreg)'s 1 / shape
      res <- -fit[[model]]$res[covariate,] / (1 / fit[[model]]$res["shape",])

@@

In the following table, we present a side-by-side comparison of the hazard ratios and confidence intervals estimated from \code{fitSmoothHazard} and the corresponding parametric model using \code{survival::survreg}, as well as the Cox model estimate. The hazard ratio estimates and confidence intervals are similar across all four models. This reinforces the idea that, under proportional hazards, we do not need to model the full hazard to obtain reliable estimates of the hazard ratio. Nevertheless, Figure \ref{fig:erspc-cox-cif} shows that different parametric models can still give rise to qualitatively different estimates for the CIF.

```{r print-erspc-estimates, eval = eval_cs1}
## ---- print-sim-table ----

kable(ptable, "latex", booktabs = TRUE, align = c("l","c","c"),
      caption = c("Comparison of estimated hazard ratios and 95\\% confidence intervals for ERSPC data."),
      col.names = c("Model","casebase::fitSmoothHazard","survival::survreg")) %>%
  kable_styling() %>%
  # add_header_above(c(" "," ","No overlap" = 2, "All causal SNPs\nin kinship" = 2, "No overlap" = 2, "All causal SNPs\nin kinship" = 2)) %>%
  # add_header_above(c(" "," ","Null model" = 4, "1% Causal SNPs" = 4)) %>%
  column_spec(1, bold = TRUE) %>%
  collapse_rows(columns = 1, latex_hline = "major", valign = "middle") %>%
  footnote(general = sprintf("Cox model estimate: HR (95%% CI) = %0.2f (%0.2f, %0.2f)", CI_cox[1], CI_cox[2], CI_cox[3]),
           general_title = "") #%>%
  # footnote(general = "Model Size is the number of selected variables using the high-dimensional BIC for \texttt{ggmix} and 10-fold cross validation for \texttt{lasso} and \texttt{twostep}.")
  #          # number = c("", "Footnote 2; "),
  #          # alphabet = c("Footnote A; ", "Footnote B; "),
  #          # symbol = c("Footnote Symbol 1; ", "Footnote Symbol 2"),
  #          general_title = "Note: ", number_title = "Type I: ",
  #          alphabet_title = "Type II: ", symbol_title = "Type III: ",
  #          footnote_as_chunk = TRUE, title_format = c("italic", "underline")
  # )
```
